---
title: "Calibration_lab"
output: html_document
date: "2026-03-01"
---

## Create an macpan SIR model spec


```{r sir_model}
library(macpan2)
library(tidyverse);theme_set(theme_bw())

## flow diagram specification
flows = list(
    mp_per_capita_flow("S", "I", "beta * I / N", "incidence")
  , mp_per_capita_flow("I", "R", "gamma", "recovery")
  , beta_thing ~ beta
)

## default values for quantities required to run simulations
default = list(
    beta = 0.2   ## transmission rate
  , gamma = 0.1  ## recovery rate
  , N = 1000      ## total population size (constant in this model)
  , I = 1        ## initial number of infectious individuals
  , R = 0        ## initial number of recovered individuals
)

## compute the initial number of susceptible individuals
initialize_state = list(S ~ N - I - R)

## model specification
spec = mp_tmb_model_spec(
    before = initialize_state
  , during = flows
  , default = default
)


```


## Deterministic and stochastic simulation of the SIR model spec


```{r sims}
# setting simulation parameters
time_steps <- 200L
outputs <- c("incidence")
nsims <- 10

# model parameter
beta <- 0.2

# simulator object
sir_simulator = mp_simulator(  
    model = spec
  , time_steps = time_steps
  , outputs = outputs
)

# simulate trajectory

## Deterministic simulation

det_sim <- (mp_trajectory(sir_simulator))


## Stochastic simulation

parameterized_sim <- (mp_tmb_calibrator(mp_euler_multinomial(spec)
	, par = "beta"
	, time = mp_sim_bounds(1,time_steps)
	, outputs = outputs
	)
)

beta_sample <- rnorm(n=nsims,mean=beta,sd=0.001)

sim_fn <- function(x){
	stochsim <- mp_trajectory_par(parameterized_sim, list(beta=x))
}

stoch_sim <- (lapply(beta_sample,sim_fn)
	|> bind_rows(.id="iter")
)

## Plotting simulations

gg <- (ggplot(det_sim,aes(x=time,y=value))
	+ geom_line()
	+ geom_line(data=stoch_sim,aes(x=time,y=value,group=iter),alpha=0.1)
)

print(gg)

simdf <- (det_sim
	|> mutate(iter = "0")
	|> bind_rows(stoch_sim)
)


```


## Fitting the model to the deterministic simulation 

```{r calibration}

## modifying the determinstic simulation as data
obsdat <- (simdf
	|> filter(iter == "0")
	|> mutate(value = round(value))
)

sir_cal = mp_tmb_calibrator(
    spec = spec
  , data = obsdat
  
  ## name the trajectory variable, with a name that
  ## is the same in both the spec and the data
  , traj = "incidence"  
  
  ## fit the following parameters
#  , par = c("beta", "gamma")
  , par = c("beta")
)

sir_opt = mp_optimize(sir_cal)

## Check optimized fit

print(sir_opt)


## Plots

fitted_data = mp_trajectory_sd(sir_cal, conf.int = TRUE)

gg <- (ggplot(obsdat)
  + geom_point(aes(time, value))
  + geom_line(data = fitted_data, aes(time, value))
  + geom_ribbon(aes(time, ymin = conf.low, ymax = conf.high)
    , data = fitted_data
    , alpha = 0.2
    , colour = "red"
  )
)

print(gg)

```

## Create a time varying SIR model with piecewise beta values

```{r timevar_model}
beta_changepoints <- c(1,50, 80, 120)
beta_values <- c(0.2,0.15,0.1,0.2)

expr <- list(beta ~ time_var(beta_values, beta_changepoints))

timevar_spec <- (spec
	|> mp_tmb_insert(phase = "during"
		, at = 1
		, expressions = expr
		, default = list(beta_values = beta_values)
		, integers = list(beta_changepoints = beta_changepoints)
	)
)

```

## Deterministic simulation of the piece-wise timevarying beta SIR
```{r timevar_sim}
# simulator object
sir_simulator = mp_simulator(  
    model = timevar_spec
  , time_steps = time_steps
  , outputs = "incidence"
)

# simulate trajectory

## Deterministic simulation

det_sim <- (mp_trajectory(sir_simulator))


gg <- (ggplot(det_sim,aes(x=time,y=value))
	+ geom_line()
)

print(gg)


```

## Fitting the piece-wise timevarying SIR to the deterministic simulation

```{r timevar_calibrate}


## modifying the determinstic simulation as data
obsdat <- (det_sim
	|> mutate(value = round(value))
)

timevar_sir_cal = mp_tmb_calibrator(
    spec = timevar_spec
  , data = obsdat
  
  ## name the trajectory variable, with a name that
  ## is the same in both the spec and the data
  , traj = "incidence"  
  
  ## fit the following parameters
#  , par = c("beta", "gamma")
  , par = c("beta_values")
)

timevar_sir_opt = mp_optimize(timevar_sir_cal)

## Check optimized fit

print(timevar_sir_opt)


## Plots

fitted_data = mp_trajectory_sd(timevar_sir_cal, conf.int = TRUE)

gg <- (ggplot(obsdat)
  + geom_point(aes(time, value))
  + geom_line(data = fitted_data, aes(time, value))
  + geom_ribbon(aes(time, ymin = conf.low, ymax = conf.high)
    , data = fitted_data
    , alpha = 0.2
    , colour = "red"
  )
)

print(gg)



```

## Starting with the original SIR model spec (model.R), fit a spline-based time varying beta SIR to the piece-wise determinsitic simulation

```{r timevar_spline_calibration}
## modifying the deterministic simulation as data


basis_cols = 5
basis_rows = time_steps
X = splines::ns(1:basis_rows
	, basis_cols
	, intercept = TRUE
	, Boundary.knots = c(1, basis_rows)
	)

timevar_spec = mp_tmb_insert_glm_timevar(spec
	, parameter_name = "beta"
	, design_matrix = X
	, timevar_coef = rep(0,
	basis_cols)
	, link_function = mp_log
)

spline_cal <- (mp_tmb_calibrator(spec = timevar_spec |> mp_rk4()
	, data = obsdat
	, time = mp_sim_bounds(1, time_steps)
	, traj = "incidence"
	, par = list(time_var_beta = mp_normal(0,log(1)))
	, outputs = c("beta_thing","incidence")
	)
)


spline_opt = mp_optimize(spline_cal)

## Check optimized fit

print(spline_opt)

## Plots

fitted_data = mp_trajectory_sd(spline_cal, conf.int = TRUE)

gg <- (ggplot(data = (fitted_data |> filter(matrix == "incidence")))
  + geom_line(aes(time, value))
  + geom_ribbon(aes(time, ymin = conf.low, ymax = conf.high)
    , alpha = 0.2
    , colour = "red"
  )
  + geom_point(data=obsdat,aes(time, value))
)

print(gg)

betaplot <- (ggplot(data = (fitted_data |> filter(matrix == "beta_thing")))
	+ geom_line(aes(time, value))
	+ geom_ribbon(aes(time, ymin=conf.low, ymax=conf.high)
		, alpha = 0.2
		, color = "red"
	)
)

print(betaplot)



```
